---
title: "Stoffel rpt2 comparison"
author: "August Arnstad"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)
showsol <- FALSE
library(knitr)
library(devtools)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 68), 
                      tidy = TRUE, 
                      warning = FALSE, 
                      error = FALSE, 
                      message = FALSE, 
                      echo = TRUE, 
                      fig.width=7, 
                      fig.height=5, 
                      fig.align="center")

library(remotes)
library(devtools)
if(!require(devtools)) install.packages("devtools")
devtools::install_github("AugustArnstad/VariableImportanceINLA")
library(VariableImportanceINLA)
```

# Comparison of the rptR method and VariableImportanceINLA
Comparison is done using examples from the rptR vignette found at:
https://cran.r-project.org/web/packages/rptR/vignettes/rptR.html
and in the article
Stoffel, M. A., Nakagawa, S., & Schielzeth, H. (2017). rptR: repeatability estimation and variance decomposition by generalized linear mixed-effects models. Methods in Ecology and Evolution, 8(11), 1639-1644.


## Compare Poisson regression (article)

```{r}
library(rptR)
library(INLA)
library(ggplot2)
library(reshape2)
library(tidyr)
library(mvtnorm)

data(BeetlesFemale)

data_pois <- BeetlesFemale

data_pois$Treatment <- as.numeric(data_pois$Treatment)

rptPois <- rptPoisson(Egg ~ Treatment+(1|Container)+(1|Population), grname=c("Container", "Population"), data=data_pois, link="log", nboot=10)


stoffel_heritability_poisson <- rptPois$R$Population[2]

poisson_formula <- Egg ~ Treatment+ f(Container, model="iid") + f(Population, model="iid") 

model_pois_glm <- VariableImportanceINLA::perform_inla_analysis(data_pois, poisson_formula, family = "poisson", link_func = "log")

pois_glm_samples <- VariableImportanceINLA::sample_posterior_count(model_pois_glm, poisson_formula, data_pois, n_samp=1000, additive_param = "Population")

```


## Compare Gaussian regression from rptR by Stoffel and VariableImportanceINLA (vignette, rep1)

```{r}
data(BeetlesBody)

data_gauss <- BeetlesBody

rpt_gauss_1 <- rpt(BodyL ~ (1 | Population), grname = "Population", data = data_gauss, datatype = "Gaussian", nboot = 10)

gauss_formula <- BodyL ~ f(Population, model="iid")

stoffel_heritability_gauss_1 <- rpt_gauss_1$R$Population
stoffel_heritability_gauss_1

model_gauss_glm <- VariableImportanceINLA::perform_inla_analysis(data_gauss, gauss_formula, family = "gaussian")

gauss_1_glm_samples <- VariableImportanceINLA::sample_posterior_gaussian(model_gauss_glm, gauss_formula, data_gauss, n_samp=1000, additive_param = "Population")

```

## Compare Gaussian with multiple random effects (vignette, rep3)

```{r}
rpt_gauss_2 <- rpt(BodyL ~ (1 | Population) + (1 | Container), grname = c("Population", "Container"), data = data_gauss, datatype = "Gaussian", nboot = 10)
    #nboot = 0, npermut = 0)

stoffel_heritability_gauss_2 <- rpt_gauss_2$R$Container

gauss_formula_multiple <- BodyL ~ f(Population, model="iid") + f(Container, model="iid")


model_gauss_glm_multiple <- VariableImportanceINLA::perform_inla_analysis(data_gauss, gauss_formula_multiple, family = "gaussian")

gauss_2_glm_samples <- VariableImportanceINLA::sample_posterior_gaussian(model_gauss_glm_multiple, gauss_formula_multiple, data_gauss, n_samp=1000, additive_param = "Container")

```

## Compare Binary regression (vignette, rep10)

```{r}
data(BeetlesMale)

data_bin <- BeetlesMale


rptBinary <- rpt(Colour ~ Treatment + (1 | Container) + (1 | Population), grname =  
    "Population", data = BeetlesMale, datatype = "Binary", nboot = 10, npermut = 0)

stoffel_heritability_binomial <- rptBinary$R$Population[2]

bin_formula <- Colour ~ Treatment + f(Container, model="iid") + f(Population, model="iid")

model_bin_glm <- VariableImportanceINLA::perform_inla_analysis(data_bin, bin_formula, family = "binomial", link_func = "logit")

bin_glm_samples <- VariableImportanceINLA::sample_posterior_count(model_bin_glm, bin_formula, data_bin, n_samp=1000, additive_param = "Population")

```

```{r}
library(dplyr)
heritability_data_poisson <- pois_glm_samples$heritability %>%
  rename(Heritability = `Heritability of: Population`)

# Create the ggplot histogram with a density overlay
heritability_poisson <- ggplot(heritability_data_poisson, aes(x = Heritability)) +
  geom_histogram(aes(y = ..density..), fill = "#C6CDF7", bins = 30, alpha = 0.7) +
  geom_density(color = "#E6A0C4", adjust = 1.5, linewidth = 1) +
  geom_vline(aes(xintercept = stoffel_heritability_poisson), color = "magenta", linetype = "dashed", size = 1) +
  labs(#title = "Distribution of Heritability of Population", 
       x = "Heritability", 
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none",
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    axis.text.x = element_text(size = 24),
    axis.text.y = element_text(size = 24)
  )

heritability_poisson

```

```{r}
library(dplyr)

names(gauss_1_glm_samples$heritability)

heritability_data_gauss_1 <- gauss_1_glm_samples$heritability %>%
  rename(Heritability = `Heritability of: Population`)

# Create the ggplot histogram with a density overlay
heritability_gauss_1 <- ggplot(heritability_data_gauss_1, aes(x = Heritability)) +
  geom_histogram(aes(y = ..density..), fill = "#C6CDF7", bins = 30, alpha = 0.7) +
  geom_density(color = "#E6A0C4", adjust = 1.5, linewidth = 1) +
  geom_vline(aes(xintercept = stoffel_heritability_gauss_1), color = "magenta", linetype = "dashed", size = 1) +
  labs(#title = "Distribution of Heritability of Population", 
       x = "Heritability", 
       y = "Density") +
  theme_minimal() + 
  theme(legend.position = "none",
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    axis.text.x = element_text(size = 24),
    axis.text.y = element_text(size = 24)
  )

heritability_gauss_1
stoffel_heritability_gauss_1
```

```{r}
library(dplyr)
heritability_data_gauss_2 <- gauss_2_glm_samples$heritability %>%
  rename(Heritability = `Heritability of: Container`)

# Create the ggplot histogram with a density overlay
heritability_gauss_2 <- ggplot(heritability_data_gauss_2, aes(x = Heritability)) +
  geom_histogram(aes(y = ..density..), fill = "#C6CDF7", bins = 30, alpha = 0.7) +
  geom_density(color = "#E6A0C4", adjust = 1.5, linewidth = 1) +
  geom_vline(aes(xintercept = stoffel_heritability_gauss_2), color = "magenta", linetype = "dashed", size = 1) +
  labs(#title = "Distribution of Heritability of Population", 
       x = "Heritability", 
       y = "Density") +
  theme_minimal() + 
  theme(legend.position = "none",
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    axis.text.x = element_text(size = 24),
    axis.text.y = element_text(size = 24)
  )

heritability_gauss_2

```


```{r}
library(dplyr)

heritability_data_binomial <- bin_glm_samples$heritability %>%
  rename(Heritability = `Heritability of: Population`)


# Create the ggplot histogram with a density overlay
heritability_binomial <- ggplot(heritability_data_binomial, aes(x = Heritability)) +
  geom_histogram(aes(y = ..density..), fill = "#C6CDF7", bins = 30, alpha = 0.7) +
  geom_density(color = "#E6A0C4", adjust = 1.5, linewidth = 1) +
  geom_vline(aes(xintercept = stoffel_heritability_binomial), color = "magenta", linetype = "dashed", size = 1) +
  labs(#title = "Distribution of Heritability of Population", 
       x = "Heritability", 
       y = "Density") +
  theme_minimal() +   
  theme(legend.position = "none",
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    axis.text.x = element_text(size = 24),
    axis.text.y = element_text(size = 24)
  )

heritability_binomial
```


```{r}
plot_folder <- "/Users/augustarnstad/Library/CloudStorage/OneDrive-NTNU/Semester_10/Master/Latex/Figures/Stoffel Comparison"

# Check if the folder exists, if not, create it
if (!dir.exists(plot_folder)) {
  dir.create(plot_folder)
}

ggsave(filename="Heritability_egg_poisson.png", plot=heritability_poisson, path=plot_folder, width = 10, height = 6, dpi = 300)
ggsave(filename="Heritability_colour_binomial.png", plot=heritability_binomial, path=plot_folder, width = 10, height = 6, dpi = 300)


```
